---
title: "Part_1_Sea_ice"
author: "Heyue WU_u7628175"
date: "15/09/2023"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

# Location on GitHub

https://github.com/AdriaWu/DS4B-short-report-1//blob/main/DS4B-short-report-1/Part_1_Sea_ice.Rmd

# 1. Loading the data_1 and preparing for the packages

```{r}
# Load necessary R packages
library(tidyverse)
library(ggplot2)
library(readxl)
library(dplyr)
library(lubridate)

# Read data from Excel file
data_1 <- read_csv("raw_data/S_seaice_extent_daily_v3.0.csv")

# have a quick view of the data
tibble(data_1)

# Check the type of value
summary(data_1)
```

# 2. Data Transformation

```{r}
# Delete the second row of data
data_1 <- data_1 |>
  slice(-1)

# Transform some columns to numbers in order to calculate or plot
data_1$Extent <- as.numeric(as.character(data_1$Extent))
data_1$Year <- as.numeric(as.character(data_1$Year))

# Check the tidy_data
tibble(data_1)
```

# 3. Filter data and evaluate by year or month

```{r}
# Filter out the data for winter
winter_data <- data_1 |>
  filter(Month %in% c(6, 7, 8))

Jun_df<- winter_data |>
  filter(Month == 6) |>
  group_by(Year) |>
  summarise(
    Mean_Jun_Extent = mean(Extent, na.rm = TRUE)
    )

July_df <- winter_data |>
  filter(Month == 7) |>
  group_by(Year) |>
  summarise(
    Mean_July_Extent = mean(Extent, na.rm = TRUE)
    )

Aug_df <- winter_data |>
  filter(Month == 8) |>
  group_by(Year) |>
  summarise(
    Mean_Aug_Extent = mean(Extent, na.rm = TRUE)
    )

Year_df <- winter_data |>
  group_by(Year) |>
  summarise(
    Min_Winter_Extent = min(Extent, na.rm = TRUE),
    Mean_Winter_Extent = mean(Extent, na.rm = TRUE),
    .groups = "drop"
  )

# Merge the data frames

# Merge Jun_df and July_df
merged_df <- left_join(Jun_df, July_df, by = "Year")

# Merge merged_df and Aug_df
merged_df <- left_join(merged_df, Aug_df, by = "Year")

# Merge merged_df and Year_df
final_df <- left_join(merged_df, Year_df, by = "Year")

# Check the results
print(Jun_df)
print(July_df)
print(Aug_df)
print(Year_df)
print(final_df)

```

# 4. Use ggplot to plot the time series

```{r}
# Emphasize the data for 2023 using points with different shape
plot_1 <- ggplot(final_df, aes(x = Year)) +
  geom_line(aes(y = Mean_Jun_Extent, color = "Mean_Jun_Extent")) +
  geom_point(data = subset(final_df, Year == 2023), aes(y = Mean_Jun_Extent), color = "green", size = 3, shape = 16) +
  
  geom_line(aes(y = Mean_July_Extent, color = "Mean_July_Extent")) +
  geom_point(data = subset(final_df, Year == 2023), aes(y = Mean_July_Extent), color = "grey", size = 3, shape = 16) +
  
  geom_line(aes(y = Mean_Aug_Extent, color = "Mean_Aug_Extent")) +
  geom_point(data = subset(final_df, Year == 2023), aes(y = Mean_Aug_Extent), color = "red", size = 3, shape = 16) +
  
  geom_line(aes(y = Min_Winter_Extent, color = "Min_Winter_Extent"), linetype = "dashed") +
  geom_point(data = subset(final_df, Year == 2023), aes(y = Min_Winter_Extent), color = "purple", size = 3, shape = 18) +
  
  geom_line(aes(y = Mean_Winter_Extent, color = "Mean_Winter_Extent"), linetype = "dashed") +
  geom_point(data = subset(final_df, Year == 2023), aes(y = Mean_Winter_Extent), color = "blue", size = 3, shape = 18) +
  
  labs(title = "Antarctic Sea Ice Extent Over the Years",
       subtitle = "Highlighting Winter 2023",
       y = "Extent (10^6 sq km)",
       color = "Legend") +
  scale_x_continuous(breaks = seq(min(final_df$Year), max(final_df$Year), by = 2)) +
  theme_minimal() +
  # Rotate x-axis label 90 degrees and center the title and subtitle
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 14))

plot_1
```

# 5. Summary the data and plot

As we journey through the visual representation of Antarctic sea ice extent over the years, a few patterns emerge. Starting from the late 1970s, we observe fluctuations in the extent of sea ice during the winter months of June, July, and August. The solid lines represent the mean extent for each of these months, while the dashed lines depict the minimum and mean extent observed during the entire winter season.

The 1980s saw a relatively stable pattern, with minor dips and rises. Moving into the 1990s, there's a noticeable increase in the extent, especially in the mid-90s. The early 2000s brought about a period of higher variability, with some years, like 2007 and 2016, showing significant drops.

However, what stands out most prominently is the data for the recent years. The year 2023, in particular, showcases a sharp decline in the extent of sea ice, reaching its lowest in the recorded history. This decline is not just a one-off event but seems to be part of a trend observed since 2019.

Such patterns raise crucial questions about the changing climate and its impact on Antarctic sea ice. While the reasons for these changes can be multifaceted, the visualization underscores the importance of continuous monitoring and research in understanding our planet's evolving climate dynamics.









