---
title: "Part_2_Frog_Species"
author: "Heyue WU_u7628175"
date: "15/09/2023"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---

# **1. Location on GitHub**

https://github.com/AdriaWu/DS4B-short-report-1//blob/main/DS4B-short-report-1/Part_2_Frog_Species.Rmd

# **2. Loading the data**

```{r}
# Load necessary R packages
library(tidyverse)
library(ggplot2)
library(readxl)
library(dplyr)
library(lubridate)

# Read data from Excel file
data_2 <- read_csv("raw_data/FrogID4_final_dataset.csv")

# have a quick view of the data
tibble(data_2)

# Check the type of value
summary(data_2)
```

# **3. Data Cleaning and Preliminary Analysis**

The next step is to identify any anomalies, inconsistencies, or missing values.

## 3.1. Check for missing values

```{r}
# Use the `summarise_all()` to summarize or calculate all columns in a data frame
data_2 |>
  summarise_all(~sum(is.na(.)))
```
This means that all rows of these columns have valid data, with no missing values. 

## 3.2. Check for duplicates 好像没太必要

```{r}
# use the `duplicated()` function to identify and count duplicate rows.
sum(duplicated(data_2))
```
So there are no duplicate rows in the data 2. 

## 3.3. Provide a summary of the dataset

Before diving into any analysis or visualization, we can make a summary of the dataset to provide a quick snapshot of key features for the data, which might help the readers who are unfamiliar with the dataset.

```{r}
# Number of unique species
unique_species <- length(unique(data_2$scientificName))

# Get range of dates

# The original date data is in character format, and we can just find the start date and end date.
start_Date <- as.character(min(data_2$eventDate))
end_Date <- as.character(max(data_2$eventDate))

# Distribution of recordings across states
state_distribution <- table(data_2$stateProvince)

# Print the summary
cat("Summary of the Dataset:\n",
    "-----------------------\n",
    "Number of Unique Species Recorded:", unique_species, "\n",
    "Range of Dates: From", start_Date, "to", end_Date, "\n",
    "Distribution of Recordings Across States:\n"
    )
print(state_distribution)
```
We can know that the dataset has recorded a total of 199 unique species. The recordings span from the date 2017-11-10 to 2021-11-09. And with 229,000 recordings, NSW dominates the dataset, accounting for a significant portion of the total recordings.

# **4. Plot the species discovery curve for Australia**

## 4.1. Plot the species discovery curve for the whole of Australia

After a preliminary analysis of the data, we can plot the species discovery curve for Australia as a whole. This will show the cumulative number of species discovered over time.

```{r}
# Convert the `eventDate` column into standardized Date objects and create a new column to store them, ensuring consistency across the dataset and better visualization later.
data_2$Date_Object <- as.Date(data_2$eventDate, format="%Y/%m/%d")

# Sort the data by date
data_sorted <- data_2 |>
  arrange(Date_Object) |>
  # Create a new column representing the cumulative number of species discovered so far, but does not include previously duplicated species.
  mutate(cumulative_species = cumsum(!duplicated(scientificName)))

# Use ggplot2 to plot the species discovery curve for Australia as whole
plot_1 <- ggplot(data_sorted, aes(x = Date_Object, y = cumulative_species)) +
  geom_line(color = "#2E86C1", linewidth = 1) +
  
  labs(title = "Species Discovery Curve for Australia",
       x = "Date",
       y = "Cumulative Number of Species Discovered") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    ) 

# Extract the first and last dates of the data_sorted 
first_date <- data_sorted[1, ]
last_date <- data_sorted[nrow(data_sorted), ]

# Add points and date labels
plot_1 <- plot_1 +
  geom_point(data = first_date, aes(x = Date_Object, y = cumulative_species), color = "#E74C3C", size = 3) +
  geom_point(data = last_date, aes(x = Date_Object, y = cumulative_species), color = "#E74C3C", size = 3) +
  geom_text(data = first_date, 
            aes(x = Date_Object, 
                y = cumulative_species,
                label = start_Date), 
            vjust = -1, 
            hjust = -0.2, 
            size = 3, 
            color = "#E74C3C") +
  geom_text(data = last_date, 
            aes(x = Date_Object, 
                y = cumulative_species, 
                label = end_Date), 
            vjust = 2, 
            hjust = 1, 
            size = 3, 
            color = "#E74C3C")

print(plot_1)

```
From plot 1, we can see that by the date 2021-11-09, the cumulative number of species discovered has already exceeded 45,000.

## 4.2. Plot the data for each state and territory

Next, let's break down the data to show the same relationship for each state and territory. For the state-wise breakdown, use different colors or line types to represent each state on the same plot. This will allow for easy comparison and identification of states with higher species diversity or more active recording.

```{r}
# Sort and group data by date and state
data_sorted_statewise <- data_2 |>
  arrange(Date_Object) |>
  # This line of code is different from the code in the plot_1
  group_by(stateProvince) |>
  mutate(cumulative_species = cumsum(!duplicated(scientificName)))

# Plot the species discovery curve for each state
plot_2 <- ggplot(data_sorted_statewise, 
                 aes(x = Date_Object, 
                     y = cumulative_species, 
                     color = stateProvince)) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Species Discovery Curve by State in Australia",
       x = "Date",
       y = "Cumulative Number of Species Discovered",
       color = "State/Territory") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 10)
  )
  
print(plot_2)
```

This plot show separate trends for each state and territory, allowing for a comparison between them. We can find that here displays a consistent trend in the data. Over time, NSW and QLD have exhibited the most significant growth in the number of species discoveries. Concurrently, the growth trends in species discoveries for WA, VIC, and NT have been relatively consistent with each other. SA, TAS, and ACT have similarly shown comparable growth patterns, while other territories have remained stable with no evident increase in species discoveries.

# **5. Estimation of True Number of Frog Species**

Now we need to estimate the true number of frog species in Australia in two ways. One is the extrapolation emthod, and another is the chang01 method.

## 5.1. Extrapolation Method

We can fit a curve (e.g., logistic growth curve) to the Australia-wide species discovery data. Extrapolate this curve to estimate when it levels off, indicating the potential total number of species.

The logistic growth curve is defined by the equation:

$$
N(t) = \frac{K}{1 + \frac{K - N_0}{N_0} e^{-rt}}
$$

Where:

- $N(t)$ is the number of species discovered up to time $t$.

- $K$ is the carrying capacity, or the total number of species.

- $r$ is the growth rate.

- $N_0$ is the initial number of species at $t = 0$.

We have sorted the data by date and has a cumulative count of species discovered up to each date before. Next, we can use the `nls()` function (nonlinear least square method) in R to fit the logistic growth curve to the data. Then use the fitted model to predict the total number of species.

```{r}
# Fit the logistic growth curve

model <- nls(cumulative_species ~ K / (1 + ((K - N0) / N0) * exp(-r * as.numeric(Date_Object - min(Date_Object)))), 
             data = data_sorted, 
             # Provide initial estimates to begin the iterative process
             start = list(K = max(data_sorted$cumulative_species) * 2, 
                          N0 = data_sorted$cumulative_species[1], 
                          r = 0.01),
             # Print details of each step in the fitting process
             trace = TRUE)

# Extract the estimated K (carrying capacity) which represents the total number of species. 解释ceof
K_estimate <- coef(model)["K"]

# Print the estimated number of species and rounding up estimates to make it easier to read.
print(paste("Estimated Total Number of Frog Species:", round(K_estimate)))
```
The output shows the iterative process of the nonlinear least squares (`nls`) method fitting the logistic growth curve. Each line represents an iteration, detailing the current sum of squared residuals (a measure of the discrepancy between the model and the data) and the estimated model parameters. Throughout the iterations, the model progressively optimized its parameter estimates. 

This is evident as the sum of squared residuals decreased from an initial `1.504456e+10` to `1.281187e+07`, indicating a gradual improvement in the model's fit to the data. The final model parameter estimates were 

- K (carrying capacity) = 195.7334

- N~0 (initial species count) = 81.10501

- r (growth rate) = 0.007125516

Based on this model, the estimated total number of frog species in Australia is **196**.

```{r}
# To visualize the fit and extrapolation
data_sorted$predicted_species <- predict(model, newdata = data_sorted)

plot_3 <- ggplot(data_sorted, aes(x = Date_Object)) +
  geom_line(aes(y = cumulative_species, color = "Observed Data")) +
  geom_line(aes(y = predicted_species, color = "Logistic Fit")) +
  labs(title = "Species Discovery Curve with Logistic Fit",
       x = "Date",
       y = "Number of Species Discovered") +
  theme_minimal() +
  scale_color_manual(values = c("Observed Data" = "blue", "Logistic Fit" = "red"))

print(plot_3)

```

## 5.2. Chao1 Method

The Chao1 estimator is a popular method to estimate species richness, especially when dealing with ecological data. 

The formula for Chao1 is:

$$
Chao1 = S_{obs} + \frac{n_1(n_1 - 1)}{2(n_2 + 1)}
$$

Where:

- $S_{obs}$ is the number of observed species.
- $n_1$ is the number of species observed exactly once (singletons).
- $n_2$ is the number of species observed exactly twice (doubletons).

The intuition behind Chao1 is that the presence of many rare species (species observed only once or twice) suggests that there are likely many species that have not been observed at all.

```{r}
# Count the number of occurrences for each species
species_counts <- table(data_2$scientificName)

# Calculate the number of singletons and doubletons
n_1 <- sum(species_counts == 1)
n_2 <- sum(species_counts == 2)

# Calculate Chao1 estimate
S_obs <- length(unique(data_2$scientificName))
Chao1 <- S_obs + (n_1 * (n_1 - 1)) / (2 * (n_2 + 1))

# For 95% confidence intervals, we use the variance formula for Chao1:
var_Chao1 <- S_obs + (n_1 * (n_1 - 1)) / (2 * (n_2 + 1))^2 * (2 * n_1 - 1) + n_1^2 / (n_2 + 1)
lower_95 <- Chao1 - 1.96 * sqrt(var_Chao1)
upper_95 <- Chao1 + 1.96 * sqrt(var_Chao1)

print(paste("Chao1 estimate:", round(Chao1)))
print(paste("95% CI: [", round(lower_95), ",", round(upper_95), "]"))

```
Using this method, we predict a total of 202 species. Furthermore, the 95% confidence interval ranges from 174 to 230. This means we are 95% confident that the true number of species lies within this range.

# **6. Comparison with Best-Guess Number**

Based on the above code and calculation, we get the following results:

- The Extrapolation Method gave us an estimate of 196 species.
- The Chao1 estimation method predicted 202 species.

Both methods provided estimates that are lower than the best-guess number.

**Discussion**

We will discuss potential reasons for discrepancies between the estimates and the best-guess number. Factors like the inherent limitations of the estimation methods, sampling bias, the geographical distribution of recordings, and seasonality can be considered.

1. **Inherent Limitations of Estimation Methods**:

- *Extrapolation Method*: This method assumes that the rate of species discovery will continue in the same pattern as observed. However, as more species are discovered, the rate of new discoveries might slow down, leading to an underestimation.

- *Chao1 Method*: This method heavily relies on the number of singletons and doubletons. If there's a high number of rarely observed species, the estimate might be skewed. Additionally, it assumes that species are randomly sampled, which might not be the case due to the reasons mentioned above.

2. **Sampling Bias**: The data might not be uniformly sampled across all regions of Australia. Some areas might be over-represented, while others might be under-represented. This can lead to an underestimation of species diversity, especially if certain regions with unique species were not adequately sampled.

3. **Geographical Distribution of Recordings**: If the recordings are predominantly from urban or easily accessible areas, many species from remote or less accessible regions might be missed. Australia's vast landscapes and diverse habitats can host a variety of species, some of which might not have been captured in the dataset.

4. **Seasonality**: Frogs are often more vocal and active during certain seasons, especially the breeding season. If recordings were predominantly from off-peak seasons, some species might not have been detected.

In conclusion, while both the Extrapolation and Chao1 methods provide valuable insights into species diversity, they have their limitations and assumptions. It's essential to consider these factors and the context of the data collection when interpreting the results. The best-guess number, derived from in-depth taxonomic and genetic studies, remains the most comprehensive estimate, but the other methods can offer quick insights when such detailed studies are not feasible.